### 2018-07-10 Manuscripts/cog-psych/revision/cogpsych.tex
to xian

The outline of the article is as follows. First we describe previous empirical work on the topic of category generation, as well as the computational formalizations of the theories in those reports. We then present two hypotheses for how contrast might affect categorization and formalize them in computational models. The first is a novel computational model, which is designed to generate categories that systematically differ from existing categories in the domain. The second assumes the goal of category generation is to create representative samples of the opposite of the observed categories. We present two experiments demonstrating strong and systematic effects of category contrast on creative generation, and we qualitatively and quantitatively analyze the performance of each model in capturing human category generation. We conclude with a discussion of the implications of our results for categorization and creative cognition, and directions for future work.

@@

\subsection{The PACKER Model}

@@

In the sections below, we present two experiments demonstrating systematic effects of category contrast on creative generation inspired by factors influencing how PACKER generates new categories. Our experiments are based on \cite{jern2013probabilistic}'s paradigm, which is a straightforward translation of the traditional artificial classification paradigm to the task of category generation: Participants are first exposed to a single, experimenter-defined category, and are then asked to generate members of a new category. We then report formal analyses comparing PACKER's account of our results to that of the hierarchical Bayesian and copy-and-tweak models developed by \cite{jern2013probabilistic}.

@@

In Experiment 1 we sought to extend our analysis of the \cite{jern2013probabilistic} data by evaluating the influence of category contrast on creative generation, given qualitatively different types of prior knowledge. We found strong evidence for effects of category contrast in each condition: Participants were more likely to generate stimuli that are more distant from (i.e., less similar to) members of a previously-learned category, and members of participant-generated categories tended to be more similar to one another than to members of previously-learned categories. We also partially replicated the classic finding that the distributional structure of generated categories reflects that of previously learned categories \citep{jern2013probabilistic,ward1994structured}: Members of generated categories were more widely distributed along dimensions which were widely distributed in the experimenter-defined category.

@@

\subsection{Representativeness in Categorization}
MAKE SURE WE CITE AND DESCRIBE \citet{thomas98}!

\subsection{Limitations and Future Directions}

Although successful in explaining our results, PACKER does not provide a full account of what is known about category generation. Most notably, in this paper we have not evaluated the model's ability to explain the classic finding that generated categories tend to share distributional commonalities with previously learned categories \citep[see][]{jern2013probabilistic,ward1994structured}. While we successfully replicated this effect in Experiment 1, we also found that its influence was limited in comparison to the fundamental constraints imposed by category contrast. Even within Experiment 1, we found systematic inconsistencies: by generating exemplars into unoccupied regions of the space, participants who learned an `XOR' category, composed of members that are widely distributed along both features and are positively correlated in space, tended to generate categories with an opposite (negative) correlation. More generally, PACKER's success over the hierarchical Bayesian model indicates that the emulation of distributional structure exerts only a limited influence in comparison to category contrast in some situations.

@@

Characterizing that problem and conducting a rational analysis is an important direction for future research. To that aim, we plan to explore the connection between exemplar modeling as an Importance Sampling approximation \citep{shi10exemplar}, and see what sort of computational-level problem PACKER approximates. Once formalized in probabilistic terms, it should also be straightforward to incorporate distributional factors into the model. Alternatively, it may be possible to integrate the core principles of PACKER into other categorization models \citep[e.g.,][]{love2004sustain,kurtz2007divergent,smith2000thirty}.

Finally, although an overall goal of this work is to advance the study of creative thought by explaining generation through empirical and quantitative approaches from research in categorization, these approaches are not easily aligned with many of the alternative theoretical viewpoints in the broader study of creativity \citep[for a review see][]{kozbelt2010theories}, such as those based on free association \citep{mednick1962associative} and conceptual combination \citep{estes2002emergence,murphy1988comprehending}. Instead, we reduced our focus by studying a highly complex behavior (creative generation) as it applies within a well-established domain (artificial category learning). We hope that future work incorporates and highlights the importance of contrast into theories of creativity.

@@

The generation of new concepts and ideas is a highly interesting topic, but it is difficult to study in a controlled experimental environment. In this paper, we have provided such an examination of category generation as it applies within an artificial categorization categorization experiment. Extending the literature on creative cognition, our experiments provide a detailed picture of the role of category contrast in generation: People seek to create concepts that are distinct from what they already know, and the nature of what is created can be influenced by what does not yet exist. Our simulations with traditional exemplar models, as well as a hierarchical Bayesian model, provide strong support for the claim that category contrast is of fundamental importance. More generally, our results demonstrate that popular explanatory approaches from basic research in cognitive science can offer a precise, quantitative account of a behavior as complex as that of creative generation.

>>>>

The outline of the article is as follows. First we describe previous empirical work on the topic of category generation, as well as the computational formalizations of the theories in those reports. We then present two hypotheses for how contrast might affect categorization and formalize them in computational models. The first is a novel computational model, which is designed to generate categories that systematically differ from existing categories in the domain. The second assumes the goal of category generation is to create representative samples of the opposite of the observed categories. We present two experiments demonstrating strong and systematic effects of category contrast on concept generation, and we qualitatively and quantitatively analyze the performance of each model in capturing human category generation. Afterwards, we find support for category contrast predicting the ease at which people learn a pair of novel categories in an empirical experiment. We conclude with a discussion of the implications of our results for categorization and creative cognition, and directions for future work.

@@

\subsection{Contrast as exemplar dissimilarity: The PACKER Model}

@@

In the sections below, we present two experiments demonstrating systematic effects of category contrast on category generation inspired by factors influencing how the contrast models generate new categories. Our experiments are based on \cite{jern2013probabilistic}'s paradigm, which is a straightforward translation of the traditional artificial classification paradigm to the task of category generation: Participants are first exposed to a single, experimenter-defined category, and are then asked to generate members of a new category. We then report formal analyses comparing the exemplar dissimilarity and the representativeness accounts of our results to that of the hierarchical Bayesian and copy-and-tweak models developed by \cite{jern2013probabilistic}. After analyzing the category generation studies, we present third experiment to demonstrate the effect of contrast in category learning. This ensures that our results are not idiosyncratic to category generation.

@@

\subsection{Discussion}
In Experiment 1 we sought to extend our analysis of the \cite{jern2013probabilistic} data by evaluating the influence of category contrast on creative generation, given qualitatively different types of prior knowledge. We found strong evidence for effects of category contrast in each condition: Participants were more likely to generate stimuli that are more distant from (i.e., less similar to) members of a previously-learned category, and members of participant-generated categories tended to be more similar to one another than to members of previously-learned categories. We also partially replicated the classic finding that the distributional structure of generated categories reflects that of previously learned categories \citep{jern2013probabilistic,ward1994structured}: Members of generated categories were more widely distributed along dimensions which were widely distributed in the experimenter-defined category.

@@

\subsection{Contrast and Representativeness in Categorization}
Xian: Please write about the two hypotheses and that followup work needs to look into why the Bayesian representativeness model doesn't do as well at explaining category learning ease, but does better at category generation.

MAKE SURE WE CITE AND DESCRIBE \citet{thomas98}!

\subsection{Limitations and Future Directions}
Although we have demonstrated that contrast plays a role in categorization, our results were equivocal as to whether contrast is governed by exemplar dissimilarity or representativeness.

Although successful in explaining our results in all three studies (albeit not as well as the representativeness model for Experiments 1 and 2), PACKER does not provide a full account of what is known about category generation. Most notably, in this paper we have not evaluated the model's ability to explain the classic finding that generated categories tend to share distributional commonalities with previously learned categories \citep[see][]{jern2013probabilistic,ward1994structured}. While we successfully replicated this effect in Experiment 1, we also found that its influence was limited in comparison to the fundamental constraints imposed by category contrast. Even within Experiment 1, we found systematic inconsistencies: by generating exemplars into unoccupied regions of the space, participants who learned an `XOR' category, composed of members that are widely distributed along both features and are positively correlated in space, tended to generate categories with an opposite (negative) correlation. More generally, PACKER inherits the strengths and weaknesses of exemplar models of categorization: It provides a simple and flexible model that explains many results, but deviates systematically from human performance in some cases.

@@

Characterizing that problem and conducting a rational analysis is an important direction for future research. To that aim, we plan to explore the connection between exemplar modeling as an Importance Sampling approximation \citep{shi10exemplar}, and see what sort of computational-level problem PACKER approximates. Once formalized in probabilistic terms, it should also be straightforward to incorporate distributional factors into the model. This would unite PACKER with the Bayesian representativeness model, and they would differ in terms of their assumptions about how people represent category distributions (as exemplars or prototypes, respectively). Alternatively, it may be possible to integrate the core principles of contrast into other categorization models \citep[e.g.,][]{love2004sustain,kurtz2007divergent,smith2000thirty}.

%Finally, although an overall goal of this work is to advance the study of creative thought by explaining generation through empirical and quantitative approaches from research in categorization, these approaches are not easily aligned with many of the alternative theoretical viewpoints in the broader study of creativity \citep[for a review see][]{kozbelt2010theories}, such as those based on free association \citep{mednick1962associative} and conceptual combination \citep{estes2002emergence,murphy1988comprehending}. Instead, we reduced our focus by studying a highly complex behavior (creative generation) as it applies within a well-established domain (artificial category learning). We hope that future work incorporates and highlights the importance of contrast into theories of creativity.

@@

The generation of new concepts and ideas is a highly interesting topic, but it is difficult to study in a controlled experimental environment. In this paper, we have provided such an examination of category generation as it applies within an artificial categorization experiment. Extending the literature on creative cognition, our experiments provide a detailed picture of the role of category contrast in generation: People seek to create concepts that are distinct from what they already know, and the nature of what is created can be influenced by what does not yet exist. Our simulations with traditional exemplar models, as well as a hierarchical Bayesian model, provide strong support for the claim that category contrast is of fundamental importance. More generally, our results demonstrate that popular explanatory approaches from basic research in cognitive science can offer a precise, quantitative account of a behavior as complex as that of creative generation.

### 2018-07-10 Manuscripts/cog-psych/revision/methods_results_e3.tex
to xian

% \begin{document}
\section{Experiment 3.}

[I'm assuming Joe will include some preamble for the experiment, but otherwise
it should be fairly straightforward for me to churn out one anyway.]

>>>>

% \begin{document}
\section{Experiment 3}
Experiments 1 and 2 clearly establish the importance of contrast in category generation. Follow-up model-based analyses illustrated that both contrast models account for participant performance better than models that do not take into account contrast. We also found more support for contrast as representativeness hypothesis over contrast as exemplar dissimilarity hypothesis.

Although category generation is an intriguing aspect of human cognition, it has played a relatively limited role in the broader categorization literature. To establish that contrast plays a broader role in categorization, we conducted a traditional category learning study. In it, participants learned two categories, where the categories learned by each participant in Experiment 3 was yoked to the Alpha shown and Beta generated by each participant in Experiment 2. If contrast plays a broader role in categorization, then Beta categories rated as "better" according to the contrast models should be easier for participants to learn.
